---
title: "Locus_QC"
format: html
editor: visual
---

# Overview and Setup

### File information

This script is for data quality control, specificially locus quality control, to ensure accuracy of subsequent analyses. Files used described below:

- `Cleaned_samples_2026.10.08` Contains all bison microsatellite data from wood and plains bison with added metadata for year collected

### Load required packages

```{r warning = FALSE, message = FALSE}
rm(list = ls())
library(adegenet)
library(PopGenReport)
library(poppr)
library(pegas)
library(strataG)
library(tidyverse)
library(graph4lg)
library(svglite)
```

### Load and prepare data

```{r}
bison.data <- read.csv("./datafiles/Cleaned_data_2026.10.08.csv",
                       check.names = F)

# Make genind object
bison.genind <- df2genind(X=bison.data[,c(15:ncol(bison.data))],
                              sep = "/",
                              ind.names = bison.data$Spreadsheet.ID,
                              loc.names = NULL,
                              pop = bison.data$Herd,
                              NA.char = "NA",
                              ploidy = 2,
                              type = "codom",
                              strata = NULL,
                              hierarchy = NULL)
bison.genind@other$year <- bison.data$Year.Collected
```

# Preliminary locus filtering

### Missing data

If a single loci contains significant missing data across samples, it could be indicative of

```{r}
locmiss_bison <-  propTyped(bison.genind, by = "loc")
locmiss_bison[which(locmiss_bison < 0.80)] # print loci with < 80% complete genotypes

barplot(locmiss_bison, ylim = c(0,1), ylab = "Complete genotypes (proportion)", xlab = "Locus", las = 2, cex.names = 0.7)

bison.genind <- missingno(bison.genind, type = "loci", cutoff = 0.20)
```

### Test for polymorphism

```{r}
summary(bison.genind)
isPoly(bison.genind) %>% summary
```

# Data subsetting

For subsequent analyses, data is divided into two temporal cohorts. Data quality control measures are run on each cohort independently as to not induce the wahlund effect.

```{r}
# Create datasets containing samples of interest for temporal comparisons
temp1 <- bison.genind[bison.genind@other$year <=2015,]

temp2 <- bison.genind[bison.genind@other$year >=2016, ]

# Remove populations which do not contain significant sample sizes in temporal groups
table(pop(temp1)) 
table(pop(temp2)) 

# Create genind summaries for both temporal groupings
temp1.sum <- summary(temp1)
temp2.sum <- summary(temp2)
```

# Hardy-Weinberg equilibrium

Because of the variance in sample sizes, and low sample sizes seen in some herds, a Markov Chain Monte Carlo exact test can be used to evaluate deviations from HWE.

### prepare data for HWE exact test

```{r}
# In order to use the genepop's MCMC method of deterimining HWE deviations, the strataG implementation can be used. the bison "genind" objects must be reformatted as "gtypes" objects
temp1.g <- genind2gtypes(temp1)
temp2.g <- genind2gtypes(temp2)

# Further, HWE deviations and linkage disequilibrium are tested within each population. Deviations or LD which occur across populations (at the same locus/loci) may indicate technical errors, adaptation, or linkage
pop_list_temp1 <- strataSplit(temp1.g)
pop_list_temp2 <- strataSplit(temp2.g)
```

### Test for HWE deviations

```{r}
# Apply HWE test from genepop. The strataG wrapper formats this back into R nicely for subsequent manipulation

# Temp1 ----
hwe_results_list_temp1 <- lapply(pop_list_temp1, function(pop) {
  hweTest(
    pop, 
    use.genepop = T,     
    which = "Proba",         
    dememorization = 10000,  
    batches = 500,           
    num.rep = 5000,
    delete.files = TRUE
  )
})

HWE.temp1 <- as.data.frame(hwe_results_list_temp1) %>% # Make results as df
  rownames_to_column("Locus") %>%  
  pivot_longer(
    cols = -c(Locus), 
    names_to = "Population",       
    values_to = "HWE_p_value"      
  ) %>% 
  # p-value correction by Population
  group_by(Population) %>% 
  mutate(p.adj_herd = p.adjust(HWE_p_value, method = "BH")) %>% 
  ungroup() 

# Temp2 ----
hwe_results_list_temp2 <- lapply(pop_list_temp2, function(pop) {
  hweTest(
    pop, 
    use.genepop = T,     
    which = "Proba",         
    dememorization = 10000,  
    batches = 500,           
    num.rep = 5000,
    delete.files = TRUE
  )
})

HWE.temp2 <- as.data.frame(hwe_results_list_temp2) %>% # Make results as df
  rownames_to_column("Locus") %>%  
  pivot_longer(
    cols = -c(Locus), 
    names_to = "Population",       
    values_to = "HWE_p_value"      
  ) %>% 
  # p-value correction by Population
  group_by(Population) %>% 
  mutate(p.adj_herd = p.adjust(HWE_p_value, method = "BH")) %>% 
  ungroup() 
```

### Summarize HWE deviations

```{r}
# Temp1 ----
temp1.HWE.sum <- HWE.temp1 %>%
  group_by(Population) %>%
  summarise(
    HWE_Deviations = sum(p.adj_herd <= 0.05, na.rm = TRUE),
  ) %>%
  arrange(desc(HWE_Deviations)) # Sort with most deviations at the top

# View the summary table
View(temp1.HWE.sum)

# Temp2 ----
temp2.HWE.sum <- HWE.temp2 %>%
  group_by(Population) %>%
  summarise(
    HWE_Deviations = sum(p.adj_herd <= 0.05, na.rm = TRUE),
  ) %>%
  arrange(desc(HWE_Deviations)) # Sort with most deviations at the top

# View the summary table
View(temp2.HWE.sum)
```

### Format significant HWE deviations

```{r}
# Temp1
HWE.sig.temp1 <- HWE.temp1 %>% 
  filter(HWE_p_value <= 0.05)

wide_HWE.temp1 <- HWE.temp1 %>%
  select(Locus, Population, p.adj_herd) %>%
  pivot_wider(
    names_from = Population, 
    values_from = p.adj_herd
  )  %>% 
  mutate(across(-Locus, ~ {
    # 1. Force 4 decimal places (turns 1 into "1.0000")
    formatted <- sprintf("%.4f", .)
    
    # 2. Add asterisk if the ORIGINAL numeric value was <= 0.05
    if_else(. <= 0.05, paste0(formatted, "*"), formatted)
  }))

write.csv(wide_HWE.temp1, "./output/tables/QC/temp1.HWE.p_adjust.csv", row.names = F)


# Temp2 
wide_HWE.temp2 <- HWE.temp2 %>%
  select(Locus, Population, p.adj_herd) %>%
  pivot_wider(
    names_from = Population, 
    values_from = p.adj_herd
  ) %>% 
  mutate(across(-Locus, ~ {
    # 1. Force 4 decimal places (turns 1 into "1.0000")
    formatted <- sprintf("%.4f", .)
    
    # 2. Add asterisk if the ORIGINAL numeric value was <= 0.05
    if_else(. <= 0.05, paste0(formatted, "*"), formatted)
  }))

write.csv(wide_HWE.temp2, "./output/tables/QC/temp2.HWE.p_adjust.csv", row.names = F)
```

# Test for linkage disequilibrium

Pairwise test for LD using Genepop through *strataG*. If LD is present in loci across the majority of herds, it may indicate physical linkage as opposed to local variation.

```{r}
# Temp1 ----
LD_results_temp1 <- lapply(pop_list_temp1, function(pop) {
  LDgenepop(
  pop,
  dememorization = 10000,
  batches = 100,
  iterations = 5000,
  delete.files = TRUE
  )
})

LD.df.temp1 <- bind_rows(LD_results_temp1, .id = "Population") # Put into dataframe format

LD.df.temp1 <- LD.df.temp1 %>%
  group_by(Population) %>%  # Separate by populations
  mutate(p.adj_herd = p.adjust(p.value, method = "BH")) %>% # Perform BH correction within each population
  ungroup()

# Dataframe of significant pairwise values
LD.temp1.sig <- LD.df.temp1 %>% 
  filter(p.adj_herd <= 0.05 & 
           !is.na(Locus.2))

write.csv(LD.temp1.sig, "./output/tables/QC/temp1LD.csv", row.names = F)


# Temp2 ----
LD_results_temp2 <- lapply(pop_list_temp2, function(pop) {
  LDgenepop(
  pop,
  dememorization = 10000,
  batches = 100,
  iterations = 5000,
  delete.files = TRUE
  )
})

LD.df.temp2 <- bind_rows(LD_results_temp2, .id = "Population") # Put into dataframe format

LD.df.temp2 <- LD.df.temp2 %>%
  group_by(Population) %>%  # Separate by populations
  mutate(p.adj_herd = p.adjust(p.value, method = "BH")) %>% # Perform BH correction within each population
  ungroup()

# Dataframe of significant pairwise values
LD.temp2.sig <- LD.df.temp2 %>% 
  filter(p.adj_herd <= 0.05 & 
           !is.na(Locus.2))

write.csv(LD.temp2.sig, "./output/tables/QC/temp2LD.csv", row.names = F)
```
